---
title: "Exploratory Data Analysis"
format: html
editor: visual
---

## Exploratory Data Analysis

First we'll run the necessary libraries for the visualizations

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
```

## Import Dataset

This dataset is about the information of used cars from CarDekho (<https://www.cardekho.com/>) collected by its creators. The dataset is provided on Kaggle \[above link\] on June 26,2020.

```{r message=FALSE, warning=FALSE}
car_df <- read_csv("CAR_DEKHO.csv")
head(car_df)
```

## Exploratory Analysis

### Summary Statistics

```{r}
summary(car_df)
```

-   Most of our variables are categorical

-   For our continuous variables there seem to be a large amount of variance (except for year) which could possibly suggest the presence of outliers.

-   Have values for 1992 to 2024

```{r}
lapply(car_df[c("fuel", "transmission", "owner")], unique)
```

-   5 categories for Fuel, 2 for Transmission and 5 for Owner

### Visualizations
