---
title: "Exploratory Data Analysis"
format: html
editor: visual
---

## Exploratory Data Analysis

First we'll run the necessary libraries for the visualizations

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(corrplot)
```

## Import Dataset

This dataset is about the information of used cars from CarDekho (<https://www.cardekho.com/>) collected by its creators. The dataset is provided on Kaggle \[above link\] on June 26,2020.

```{r message=FALSE, warning=FALSE}
car_df <- read_csv("CAR_DEKHO.csv")
head(car_df)
```

## Exploratory Analysis

### Summary Statistics

```{r message=FALSE, warning=FALSE}
str(car_df) 
```

-   We count with 5 character and 3 numerical variables

```{r}
summary(car_df)
```

-   Most of our variables are categorical

-   For our continuous variables there seem to be a large amount of variance (except for year) which could possibly suggest the presence of outliers.

-   Have values for 1992 to 2024

```{r}
lapply(car_df[c("fuel", "transmission", "owner")], unique)
```

-   5 categories for Fuel, 2 for Transmission and 5 for Owner

### Missing Values

```{r}
colSums(is.na(car_df))  
sum(is.na(car_df))    
```

-   No missing values so that means we are not going to need to perform transformations

### Visualizations

```{r}
# Plot for 'year'
ggplot(car_df, aes(x = year)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(title = "Distribution of Year", x = "Year", y = "Count") +
  theme_minimal()

# Plot for 'selling_price'
ggplot(car_df, aes(x = selling_price)) +
  geom_histogram(binwidth = 50000, fill = "green", color = "white") +
  labs(title = "Distribution of Selling Price", x = "Selling Price (in currency units)", y = "Count") +
  theme_minimal()

# Plot for 'km_driven'
ggplot(car_df, aes(x = km_driven)) +
  geom_histogram(binwidth = 10000, fill = "purple", color = "white") +
  labs(title = "Distribution of KM Driven", x = "Kilometers Driven", y = "Count") +
  theme_minimal()
```

-   We don't seem to count with normal distributions for neither of the 3 variables

-   Year seems to have a left skew distribution

-   Selling Price and KM Driven show a right-skewed distribution, probably hinting on outliers for high values

```{r}
car_df$owner <- factor(car_df$owner, 
                       levels = c("First Owner", 
                                  "Second Owner", 
                                  "Third Owner", 
                                  "Fourth & Above Owner", 
                                  "Test Drive Car"))

# Bar plot with the specified order
ggplot(car_df, aes(x = owner)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Owner Variable",
       x = "Owner",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Smaller and angled text
        plot.title = element_text(size = 14))
```

-   The distribution for the owner variable shows a intuitive pattern where most of the cars offered have just one owner, and then the proportion decreases by almost half for each new owner that has had the vehicle.

```{r}
# Box plot for selling price by fuel type
ggplot(car_df, aes(x = fuel, y = selling_price, fill = fuel)) +
  geom_boxplot(outlier.shape = 16, outlier.size = 2) +
  labs(title = "Box Plot of Selling Price by Fuel Type", 
       x = "Fuel Type", 
       y = "Selling Price (in currency units)") +
  theme_minimal() +
  theme(legend.position = "none") 
```

-   Outliers seem to be making hard the interpretation for difference in selling price depending on the type of Fuel

-   Diesel and Petrol seem to have most of the outliers but could be because of the size of the data for each type

```{r message=FALSE, warning=FALSE}

lower_bound <- quantile(car_df$selling_price, 0.05, na.rm = TRUE)  
upper_bound <- quantile(car_df$selling_price, 0.95, na.rm = TRUE)  

# Box plot for selling price by fuel type (without outliers)
ggplot(car_df, aes(x = fuel, y = selling_price, fill = fuel)) +
  geom_boxplot(outlier.shape = NA) +
  labs(title = "Box Plot of Selling Price by Fuel Type (Without Outliers)", 
       x = "Fuel Type", 
       y = "Selling Price (in currency units)") +
  scale_y_continuous(limits = c(lower_bound, upper_bound)) +
  theme_minimal() +
  theme(legend.position = "none")
```

-   Without outliers we can see better that there could potentially be a difference in the mean selling price depending on the fuel type

-   Also, we can better understand the distribution of the selling price for each fuel type

```{r}
ggplot(car_df, aes(x = year, y = selling_price)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "Scatter Plot of Selling Price vs. Year",
       x = "Year",
       y = "Selling Price (in currency units)") +
  theme_minimal()
```

-   For year and selling price, even though we don't have a straight away linear regression, we can see that the overall values seem to increase as the year go buy, which is intuitive since older cars always tend to have lower prices.

```{r}
ggplot(car_df, aes(x = year, y = selling_price)) +
  geom_point(aes(color = transmission), alpha = 0.6) +
  labs(title = "Scatter Plot of Selling Price vs. Year by Transmission Type",
       x = "Year",
       y = "Selling Price (in currency units)",
       color = "Transmission") +
  facet_wrap(~ transmission) +  # Create separate plots for each transmission type
  theme_minimal() +
  theme(legend.position = "none")
```

-   When we add the transmission variable to this plot, we can see that the analysis changes, since now we can see a more linear relationship for Manual cars, when as for Automatic there seems to be more variance and a potential higher slope

```{r}
num_vars <- car_df[sapply(car_df, is.numeric)]

cor_matrix <- cor(num_vars)

corrplot(cor_matrix, method = 'color', order = 'alphabet', tl.col = "black", tl.srt = 45)
```

-   The correlation matrix supports our hypothesis that newer cars (higher year value) tend to be more expensive. This is reflected in a positive correlation with selling price, though it is not as strong as expected.

-   Kilometers driven have a negative correlation with selling price, meaning fewer kilometers result in higher prices. This aligns with the logic that a car with low mileage is closer to being new and therefore more valuable.

-   The correlation between year and km_driven shows that newer cars are less likely to have high mileage, as older cars have been in use longer. This results in a significant negative correlation.
